Towards Practical First-Order Model Counting

### 2024-04-22 doc/paper.tex
Section 4.1

\todo[inline,caption={}]{
  \begin{itemize}
    \item 9 pages!!!
    \item Add some papers mentioned in:
          \begin{itemize}
            \item very recent work
            \item my previous paper, including:
                  \begin{itemize}
                    \item other liftable fragments
                    \item some more theory papers, e.g., LICS 2018
                  \end{itemize}
          \end{itemize}
  \end{itemize}
}

@@

\begin{figure*}[t]
  \centering
  \begin{tikzpicture}
    \node at (-1, 0) (formula) {$\phi$};
    \node[draw,rounded rectangle] at (3, 0) (alg1) {\CompileWithBaseCases};
    \node[draw,rounded rectangle] at (9, 0) (compilation) {Compile to C++};

@@

    \node[draw,rounded rectangle] at (3, -2) (alg2) {\FindBaseCases};
    \node[draw,rounded rectangle,left = 0.1cm of alg2] (crane) {\Crane};
    \node[draw,rounded rectangle,right = 0.1cm of alg2] (alg3) {\Propagate};
    \node[draw,rounded rectangle,right = 0.1cm of alg3] (simplify) {\Simplify};

    \node[draw,fit={(alg1) (compilation) (crane) (alg2) (alg3)},inner ysep=7pt,yshift=5pt] {};
    \node at (0.2, 0.5) {\Cranetwo};

    \draw[-Latex] (formula) -- (alg1);
    \draw[-Latex] (alg1) -- node[above] {$\mathcal{E}$} (compilation);
    \draw[-Latex] (compilation) -- (cpp);
    \draw[-Latex] (sizes) -- (cpp);
    \draw[-Latex] (cpp) -- (count);

    \draw[-Latex,dashed] (alg1) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (alg1) -- node[midway,left] {uses} (alg2);
    \draw[-Latex,dashed] (alg1) -- node[midway,left] {uses} (alg3);
    \draw[-Latex,dashed] (alg1) -- node[midway,right] {uses} (simplify);
  \end{tikzpicture}
  \caption[]{The outline of using \Cranetwo{} to compute the model count of a
    formula $\phi$. The formula is compiled into a set of equations
    $\mathcal{E}$ that are then compiled to a C++ program. This program can then
    be run with different command line arguments to compute the model count of
    $\phi$ for various domain sizes. \CompileWithBaseCases makes use
    of: \begin{enumerate*}[label=(\roman*)]
      \item the knowledge compilation algorithm of \textsc{Crane} (denoted by \Crane),
      \item a procedure for identifying a sufficient set of base cases (denoted by \FindBaseCases),
      \item a procedure for constructing a formula that corresponds to a given base case (denoted by \Propagate), and
      \item algebraic simplification techniques (denoted by \Simplify).
    \end{enumerate*}
  }\label{fig:overview}
\end{figure*}

\todo[inline]{By pointing to the outline in \cref{fig:overview}, make forward
  references to the sections of the paper.}

@@

\begin{example}\label{example:functions}
  Functions
  \begin{itemize}
    \item In \Ctwo: $\forall x \in \Delta\text{.
          }\exists^{=1} y \in \Delta\text{. }P(x, y)$
    \item In \UFO:
          \begin{gather*}
            \forall x, y \in \Delta\text{. }S(x) \lor \neg P(x, y)\\
            |P| = |\Delta|
          \end{gather*}
    \item In \FO:
          \begin{gather*}
            \forall x \in \Delta\text{. }\exists y \in \Delta\text{. }P(x, y)\\
            \forall x, y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z
          \end{gather*}
  \end{itemize}
\end{example}

@@

\begin{algorithm}[t]
  \caption{\protect\CompileWithBaseCases{$\phi$}}\label{alg:main}
  \KwIn{formula $\phi$}
  \KwOut{set $\mathcal{E}$ of equations}
  $(\mathcal{E}, \mathcal{F}, \mathcal{D}) \gets \Crane{$\phi$}$\;\label{line:first}
  $\mathcal{E} \gets \Simplify{$\mathcal{E}$}$\;\label{line:second}
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$\mathcal{E}$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{$i$ such that $x_{i} \in \mathbb{N}_{0}$}{
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $\mathcal{E} \gets \mathcal{E} \cup \CompileWithBaseCases{$\psi$}$\;
  }
\end{algorithm}

\Cref{alg:main} outlines our overall approach for compiling a formula into a set
of equation that include the required base cases. In short, we first use the
knowledge compilation algorithm of the original
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23} to compile the formula into:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
After some algebraic simplification, $\mathcal{E}$ is passed to the
\FindBaseCases procedure (described in \cref{sec:identifying}) that returns a
set of base cases that we need to find solutions for. For each base case
$f(\mathbf{x})$, we identify the formula associated with $f$ and simplify it
using the \Propagate procedure (described in \cref{sec:simplifying}). The
algorithm then recurses on these simplified formulas and adds the resulting base
case equations to $\mathcal{E}$. \Cref{example:overall} explains \cref{alg:main}
in more detail.

@@

\begin{algorithm}[t]
  \caption{\protect\FindBaseCases{$\mathcal{E}$}}\label{alg1}
  \KwIn{set $\mathcal{E}$ of equations}
  \KwOut{set $\mathcal{B}$ of base cases}

@@

\Cref{alg1} summarises the implementation of \FindBaseCases. For each recursive
call from a function $f$ to itself, we consider two types of arguments:
\begin{enumerate*}[label=(\roman*)]
  \item constants and
  \item arguments of the form $x_{i} - c_{i}$, where $c_{i} \in \mathbb{N}_{0}$
  is a constant, and $x_{i}$ is the $i$-th argument of the signature of $f$.
\end{enumerate*}
In the former case, we consider a base case specifically for that constant. In
the latter case, we consider a base case for constants from zero up to (but not
including) $c_{i}$. \Cref{thm:halting} below motivates this approach.

@@

\begin{theorem}[Halting]\label{thm:halting}
  Let us assume the following about the input set of equations $\mathcal{E}$:
  \begin{enumerate}
    \item Each function $f$ has exactly one equation with $f$ on the left-hand
          side.\@ We call this equation the \emph{definition} of $f$.
    \item There exists a topological ordering of all functions ${(f_{i})}_{i}$
          such that the definition of $f_{i}$ does not contain function calls to
          $f_{j}$ with $j > i$.\footnote{This condition excludes the possibility
          of mutual recursion and similar cyclic scenarios and is akin to
          stratified logic programs~\cite{DBLP:books/sp/Lloyd87}.}
    \item For every equation $(f(\mathbf{x}) = \expr) \in \mathcal{E}$, every
          recursive function call $f(\mathbf{y}) \in \expr$ satisfies the
          following:
          \begin{itemize}
            \item each $y_{i}$ is either $x_{i} - c_{i}$ or $c_{i}$ for some
                  constant $c_{i} \in \mathbb{N}_{0}$;
            \item there exists $i$ such that $y_{i} = x_{i} - c_{i}$ for some
                  $c_{i} > 0$.
          \end{itemize}
  \end{enumerate}
  Then the set of base cases $\mathcal{B}$ returned by
  \FindBaseCases{$\mathcal{E}$} is \emph{sufficient} for $\mathcal{E}$ in the
  following sense. The evaluation of any function call with non-negative
  arguments halts as long as the evaluation of base cases is prioritised over
  the corresponding recursive definitions.\footnote{Recall that, as previously
    discussed, the order in which base cases are considered is immaterial.}
\end{theorem}

@@

\begin{algorithm}[t]
  \caption{\protect\Propagate{$\phi$, $\Delta$, $n$}}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, $n \in \mathbb{N}_{0}$}
  \KwOut{formula $\phi'$}
  $\phi' \gets \emptyset$\;
  \uIf{$n = 0$}{
    \ForEach{clause $C \in \phi$}{
      \lIf{$\Delta \not\in \Doms(C)$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      \Else{
        $C' \gets \{\, l \in C \mid \Delta \not\in \Doms(l) \,\}$\;
        \If{$C' \ne \emptyset$}{
          $l \gets \text{an arbitrary literal in } C'$\;\label{line:smoothing1}
          $\phi' \gets \phi' \cup \{\, C' \cup \{\, \neg l \,\} \,\} $\;\label{line:smoothing2}
        }
      }
    }
  }
  \Else{
    $D \gets \text{a set of $n$ new constants in $\Delta$}$\;
    \ForEach{clause $C \in \phi$}{
      ${(x_{i})}_{i=1}^{m} \gets \text{the variables in $C$ with domain $\Delta$}$\;
      \lIf{$m = 0$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      \Else{
        $\phi' \gets \phi' \cup \{\, C[x_{1} \mapsto c_{1}, \dots, x_{m} \mapsto c_{m}] \mid {(c_{i})}_{i=1}^{m} \in D^{m} \,\}$\;
      }
    }
  }
\end{algorithm}

\Propagate (\cref{alg2}) modifies formula $\phi$ with the assumption that domain
$\Delta$ has size $n \in \mathbb{N}_{0}$. In the case of $n=0$, many clauses
become vacuously satisfied and can be removed. In the case of
$n > 0$\footnote{None of the formulas considered in this work had $n > 1$.}, we
perform partial grounding, using constants to replace all variables quantified
over $\Delta$. \Cref{alg2} considers these two cases separately. For a literal
or a clause $C$, we write $\Doms(C)$ to denote the set of corresponding domains.

@@

\subsection{Smoothing for Base Cases (TODO)}\label{sec:smoothingbase}

\todo[inline]{Must include an example of propagating the empty-domain
  assumption (ideally based on previous examples).}

\paragraph{Key point.}
If there is a predicate $P$ that has nothing to do with domain $\Delta$, make
sure that $P$ remains as part of the formula even if all clauses with $P$ can be
removed. Which literal is picked on \cref{line:smoothing1} of \cref{alg2} is not
important because any choice achieves the same goal: constructing a clause that
mentions the same predicates as $C'$ and is satisfied by any structure.

\begin{fact}
  Assuming that domain $\Delta$ is empty, any clause that contains
  `$\forall x \in \Delta$' (for any variable $x$) is vacuously satisfied by all
  structures.
\end{fact}

For example, consider the formula
\begin{gather}
  \forall x \in \Delta\text{. }\forall y, z \in \Gamma\text{. }P(x) \lor Q(y, z)\label[clause]{eq:example1}\\
  \forall y, z \in \Gamma'\text{. }Q(y, z)\label[clause]{eq:example2}
\end{gather}
and assume that $\Gamma' \subseteq \Gamma$. If we set $|\Delta|$
to zero and remove clauses with variables quantified over $\Delta$, we get
\begin{equation}\label[clause]{eq:simplified}
  \forall y, z \in \Gamma'\text{. }Q(y, z),
\end{equation}
but the model count of \cref{eq:simplified} is one. However, the actual model
count should be $2^{|\Gamma|^2 - |\Gamma'|^2}$. That is, $Q$ as a relation is a
subset of $\Gamma \times \Gamma$. While \cref{eq:example1} becomes vacuously
true, \cref{eq:example2} fixes the value of $Q$ over
$\Gamma' \times \Gamma' \subseteq \Gamma \times \Gamma$. Hence, the number of
different values that $Q$ can take is
$|(\Gamma \times \Gamma) \setminus (\Gamma' \times \Gamma')| = |\Gamma|^{2} - |\Gamma'|^{2}$.

We address this issue by converting clauses with universal quantifiers over the
empty domain to tautologies, hence retaining all the predicates that have no
argument assigned to the empty domain. For example, we would convert
\cref{eq:example1,eq:example2} to
\begin{gather*}
  \forall y, z \in \Gamma\text{. }Q(y, z) \lor \neg Q(y, z)\\
  \forall y, z \in \Gamma'\text{. }Q(y, z).
\end{gather*}
The model count returned by this will also consider the truth value of $Q$ over
$y \in \Gamma \setminus \Gamma'$ or $z \in \Gamma \setminus \Gamma'$.

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item It seems like I'll need to introduce:
          \begin{itemize}
            \item FCGs
            \item node types (be consistent w.r.t.\ node vs vertex)
            \item their notation
          \end{itemize}
    \item Describe \cref{fig:smoothing1,fig:smoothing2}
  \end{itemize}
}

\begin{figure}
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\DR(c \in \Delta)$,name=dr,ellipse,draw,label={right:\textcolor{blue}{3. $\{\, \forall x \in \Delta\text{. }P(x) \,\}$}}
    [$\land$,ellipse,draw,label={left:\textcolor{blue}{2. $\{\, P(c) \,\}$}},label={right:\textcolor{blue}{6. $\{\, P(c), \forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \,\}$}},
    [$P(c)$,rectangle,draw,fill=green!20,label={272:\textcolor{blue}{1. $\{\, P(c) \,\}$}}],
    [$\CR(\Delta' \gets \Delta \setminus \{\, c \,\})$,ellipse,draw,label={87:\textcolor{blue}{5. $\{\, \forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \,\}$}}
    [$\Reff(\Delta \mapsto \Delta')$,name=ref,ellipse,draw,label={below:\textcolor{blue}{4. $\{\, \forall x \in \Delta'\text{. }P(x) \,\}$}}]
    ]
    ]
    ]
  \end{forest}
  \caption{An example based on permutation counting (but simplified a lot). The
    arc from the $\Reff$ node to the $\DR$ node is omitted.}\label{fig:smoothing1}
\end{figure}

\begin{figure}
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\DR(c \in \Delta)$,name=dr,ellipse,draw,
    [$P(c)$,rectangle,draw,fill=green!20],
    [$\forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \lor \neg P(x)$,rectangle,dashed,edge=dashed,draw,fill=blue!20]
    ]
  \end{forest}
  \caption{An artificial example of a situation when smoothing nodes need to be
    added below a $\DR$ node}\label{fig:smoothing2}
\end{figure}

\begin{algorithm}
  \caption{Propagate atoms for smoothing across the FCG}\label{alg:smoothing}
  \SetKwData{changed}{changed}
  \KwIn{FCG $(V, s, N^+, \tau)$}
  \KwIn{function $\iota$ that maps vertex types in $\mathcal{T}$ to sets of atoms}
  \KwIn{functions $\{\,f_t\,\}_{t \in \mathcal{T}}$ that map a list of sets of atoms to a set of atoms}
  \KwOut{function $S$ that maps vertices in $V$ to sets of atoms}
  $S \gets \{\, v \mapsto \iota(\tau(v)) \mid v \in V \,\}$\;
  $\changed \gets \texttt{true}$\;
  \While{\changed}{
    $\changed \gets \texttt{false}$\;
    \ForEach{vertex $v \in V$}{
      $S' \gets f_{\tau(v)}(\langle S(w) \mid w \in N^+(v) \rangle)$\;
      \If{$S' \ne S(v)$}{
        $\changed \gets \texttt{true}$\;
        $S(v) \gets S'$\;
      }
    }
  }
\end{algorithm}

Smoothing is a two-step process. First, atoms that are still accounted for in
the circuit are propagated upwards. Then, at vertices of certain types, missing
atoms are detected and additional sinks are created to account for them. If left
unchanged, the first step of this process would result in an infinite loop
whenever a cycle is encountered. \Cref{alg:smoothing} outlines how the first
step can be adapted to an arbitrary directed graph.

@@

\begin{description}
  \item[\texttt{Ref}.] During smoothing, when unit clauses are propagated in the
        opposite direction of the FCG arcs (i.e., `upwards'), when visiting a
        \texttt{Ref} node, these clauses are translated using the domain map of
        the \texttt{Ref} node. For example, if the domain map include
        $\Delta \mapsto \Delta'$, and the unit clause mentions $\Delta$, then
        replace it by $\Delta'$.
  \item[Constraint removal.] Do reverse constraint removal. Assuming that
        domain $\Delta$ with constraints $x \ne c$ (for some constant
        $c \in \Delta$) were replaced with domain $\Delta'$, replace each
        $\forall x \in \Delta'\text{. }\phi(x)$ with
        $\forall x \in \Delta\text{. }X \ne c \Rightarrow \phi(x)$.
  \item[Domain recursion.] Suppose the domain recursion node introduces constant
        $c \in \Delta$. For each unit clause received from the child node,
        replace each occurrence of $\phi(x)$ or $\forall x \in \Delta\text{.
        } x \ne c \Rightarrow \phi(x)$ with $\forall x \in \Delta\text{.
        }\phi(x)$. This can be seen as a claim about what ground atoms the
        domain recursion node \emph{should} cover (or a temporary assumption).
        If the relevant subgraph indeed covers those ground atoms, Stage~2 will
        do nothing. Otherwise, smoothing nodes will be added below the domain
        recursion node to cover the difference between what was propagated from
        the domain recursion node and what was received from the child node.
\end{description}

@@

\todo[inline]{Somewhere: mention that we're primarily concerned with computing
  unweighted model counts, but everything trivially extends to the weighted
  setting as well.}

>>>>

\todo[inline,caption={}]{
  \begin{itemize}
    \item 9 pages including appendices but not acknowledgements
    \item Add some papers mentioned in:
          \begin{itemize}
            \item very recent work
            \item my previous paper, including:
                  \begin{itemize}
                    \item other liftable fragments
                    \item some more theory papers, e.g., LICS 2018
                  \end{itemize}
          \end{itemize}
  \end{itemize}
}

@@

\begin{figure*}[t]
  \centering
  \begin{tikzpicture}
    \node at (-1, 0) (formula) {$\phi$};
    \node[draw,rounded rectangle] at (3, 0) (compilewithbasecases) {\CompileWithBaseCases};
    \node[draw,rounded rectangle] at (9, 0) (compilation) {Compile to C++};

@@

    \node[draw,rounded rectangle] at (3, -2) (findbasecases) {\FindBaseCases};
    \node[draw,rounded rectangle,left = 0.1cm of findbasecases] (crane) {\Crane};
    \node[draw,rounded rectangle,right = 0.1cm of findbasecases] (propagate) {\Propagate};
    \node[draw,rounded rectangle,right = 0.1cm of propagate] (simplify) {\Simplify};

    \node[draw,fit={(compilewithbasecases) (compilation) (crane) (findbasecases) (propagate)},inner ysep=7pt,yshift=5pt] {};
    \node at (0.2, 0.5) {\Cranetwo};

    \draw[-Latex] (formula) -- (compilewithbasecases);
    \draw[-Latex] (compilewithbasecases) -- node[above] {$\mathcal{E}$} (compilation);
    \draw[-Latex] (compilation) -- (cpp);
    \draw[-Latex] (sizes) -- (cpp);
    \draw[-Latex] (cpp) -- (count);

    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (findbasecases);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (propagate);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,right] {uses} (simplify);
  \end{tikzpicture}
  \caption[]{The outline of using \Cranetwo{} to compute the model count of a
    formula $\phi$. The formula is compiled into a set of equations
    $\mathcal{E}$ that are then compiled to a C++ program. This program can then
    be run with different command line arguments to compute the model count of
    $\phi$ for various domain sizes. \CompileWithBaseCases makes use
    of: \begin{enumerate*}[label=(\roman*)]
      \item the knowledge compilation algorithm of \textsc{Crane} (denoted by \Crane),
      \item a procedure for identifying a sufficient set of base cases (denoted by \FindBaseCases),
      \item a procedure for constructing a formula that corresponds to a given base case (denoted by \Propagate), and
      \item algebraic simplification techniques (denoted by \Simplify).
    \end{enumerate*}
  }\label{fig:overview}
\end{figure*}

\todo[inline,caption={}]{
  \begin{itemize}
    \item By pointing to the outline in \cref{fig:overview}, make forward
          references to the sections of the paper.
    \item Somewhere: mention that we're primarily concerned with computing
          unweighted model counts, but everything trivially extends to the
          weighted setting as well.
    \item Introduce greedy search and breadth-first search (BFS), together with
          \Cranegreedy{} and \Cranebfs{}. The knowledge compilation part of both
          \textsc{Crane} and \Cranetwo{} can be executed using either greedy
          (similar to \textsc{ForcLift}) or breadth-first search. We use both in
          our experiments, denoting them as \Cranegreedy{} and \Cranebfs{},
          respectively.
    \item Introduce the WFOMC acronym.
    \item Introduce first-order knowledge compilation as a class of WFOMC
          algorithms, defining/mentioning compilation rules, FCG, etc.
  \end{itemize}
}

@@

\begin{example}\label{example:functions}
  Functions
  \begin{itemize}
    \item In \Ctwo: $\forall x \in \Delta\text{.
          }\exists^{=1} y \in \Delta\text{. }P(x, y)$
    \item In \UFO:
          \begin{gather*}
            \forall x, y \in \Delta\text{. }S(x) \lor \neg P(x, y)\\
            |P| = |\Delta|
          \end{gather*}
    \item In \FO:
    \begin{equation}\label[formula]{eq:fo}
      \begin{gathered}
        \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
        \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z
      \end{gathered}
    \end{equation}
  \end{itemize}

  Note that \cref{eq:fo} is written with two domains instead of one. Clearly,
  the model count does not change whenever $|\Gamma| = |\Delta|$. As was
  previously observed by Dilkas and Belle~\shortcite{DBLP:conf/kr/DilkasB23},
  such a reformulation can help first-order knowledge compilation algorithms
  find efficient solutions.
\end{example}

@@

\begin{algorithm}[t]
  \caption{\protect\CompileWithBaseCases{$\phi$}}\label{alg:compilewithbasecases}
  \KwIn{formula $\phi$}
  \KwOut{set $\mathcal{E}$ of equations}
  $(\mathcal{E}, \mathcal{F}, \mathcal{D}) \gets \Crane{$\phi$}$\;\label{line:first}
  $\mathcal{E} \gets \Simplify{$\mathcal{E}$}$\;\label{line:second}
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$\mathcal{E}$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{$i$ such that $x_{i} \in \mathbb{N}_{0}$}{
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $\mathcal{E} \gets \mathcal{E} \cup \CompileWithBaseCases{$\psi$}$\;
  }
\end{algorithm}

\Cref{alg:compilewithbasecases} outlines our overall approach for compiling a
formula into a set of equation that include the required base cases. In short,
we first use the knowledge compilation algorithm of the original
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23} to compile the formula into:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
After some algebraic simplification, $\mathcal{E}$ is passed to the
\FindBaseCases procedure (described in \cref{sec:identifying}) that returns a
set of base cases that we need to find solutions for. For each base case
$f(\mathbf{x})$, we identify the formula associated with $f$ and simplify it
using the \Propagate procedure (described in \cref{sec:simplifying}). The
algorithm then recurses on these simplified formulas and adds the resulting base
case equations to $\mathcal{E}$. \Cref{example:overall} explains
\cref{alg:compilewithbasecases} in more detail.

@@

\begin{algorithm}[t]
  \caption{\protect\FindBaseCases{$\mathcal{E}$}}\label{alg:findbasecases}
  \KwIn{set $\mathcal{E}$ of equations}
  \KwOut{set $\mathcal{B}$ of base cases}

@@

\Cref{alg:findbasecases} summarises the implementation of \FindBaseCases. For
each recursive call from a function $f$ to itself, we consider two types of
arguments:
\begin{enumerate*}[label=(\roman*)]
  \item constants and
  \item arguments of the form $x_{i} - c_{i}$, where $c_{i} \in \mathbb{N}_{0}$
  is a constant, and $x_{i}$ is the $i$-th argument of the signature of $f$.
\end{enumerate*}
In the former case, we consider a base case specifically for that constant. In
the latter case, we consider a base case for constants from zero up to (but not
including) $c_{i}$. \Cref{thm:halting} below motivates this approach.

@@

\begin{theorem}[Termination]\label{thm:halting}
  Let us assume the following about the input set of equations $\mathcal{E}$:
  \begin{enumerate}
    \item Each function $f$ has exactly one equation with $f$ on the left-hand
          side.\@ We call this equation the \emph{definition} of $f$.
    \item There exists a topological ordering of all functions ${(f_{i})}_{i}$
          such that the definition of $f_{i}$ does not contain function calls to
          $f_{j}$ with $j > i$.\footnote{This condition excludes the possibility
          of mutual recursion and similar cyclic scenarios and is akin to
          stratified logic programs~\cite{DBLP:books/sp/Lloyd87}.}
    \item For every equation $(f(\mathbf{x}) = \expr) \in \mathcal{E}$, every
          recursive function call $f(\mathbf{y}) \in \expr$ satisfies the
          following:
          \begin{itemize}
            \item each $y_{i}$ is either $x_{i} - c_{i}$ or $c_{i}$ for some
                  constant $c_{i} \in \mathbb{N}_{0}$;
            \item there exists $i$ such that $y_{i} = x_{i} - c_{i}$ for some
                  $c_{i} > 0$.
          \end{itemize}
  \end{enumerate}
  Then the set of base cases $\mathcal{B}$ returned by
  \FindBaseCases{$\mathcal{E}$} is \emph{sufficient} for $\mathcal{E}$ in the
  following sense. The evaluation of any function call with non-negative
  arguments terminates as long as the evaluation of base cases is prioritised
  over the corresponding recursive definitions.\footnote{Recall that, as
    previously discussed, the order in which base cases are considered is
    immaterial.}
\end{theorem}

@@

\begin{algorithm}[t]
  \caption{\protect\Propagate{$\phi$, $\Delta$, $n$}}\label{alg:propagate}
  \KwIn{formula $\phi$, domain $\Delta$, $n \in \mathbb{N}_{0}$}
  \KwOut{formula $\phi'$}
  $\phi' \gets \emptyset$\;
  \uIf{$n = 0$}{
    \ForEach{clause $C \in \phi$}{
      \lIf{$\Delta \not\in \Doms(C)$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      \Else{
        $C' \gets \{\, l \in C \mid \Delta \not\in \Doms(l) \,\}$\;
        \If{$C' \ne \emptyset$}{
          $l \gets \text{an arbitrary literal in } C'$\;\label{line:smoothing1}
          $\phi' \gets \phi' \cup \{\, C' \cup \{\, \neg l \,\} \,\} $\;\label{line:smoothing2}
        }
      }
    }
  }
  \Else{
    $D \gets \text{a set of $n$ new constants in $\Delta$}$\;
    \ForEach{clause $C \in \phi$}{
      ${(x_{i})}_{i=1}^{m} \gets \text{the variables in $C$ with domain $\Delta$}$\;
      \lIf{$m = 0$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      \Else{
        $\phi' \gets \phi' \cup \{\, C[x_{1} \mapsto c_{1}, \dots, x_{m} \mapsto c_{m}] \mid {(c_{i})}_{i=1}^{m} \in D^{m} \,\}$\;
      }
    }
  }
\end{algorithm}

\Propagate (\cref{alg:propagate}) modifies formula $\phi$ with the assumption
that domain $\Delta$ has size $n \in \mathbb{N}_{0}$. In the case of $n=0$, many
clauses become vacuously satisfied and can be removed. In the case of
$n > 0$\footnote{None of the formulas considered in this work had $n > 1$.}, we
perform partial grounding, using constants to replace all variables quantified
over $\Delta$. \Cref{alg:propagate} considers these two cases separately. For a
literal or a clause $C$, we write $\Doms(C)$ to denote the set of corresponding
domains.

@@

\subsection{Smoothing for Base Cases}\label{sec:smoothingbase}

In this section, we motivate and describe \cref{line:smoothing1,line:smoothing2}
of \cref{alg:propagate}. Suppose that \Propagate is called with arguments
$(\phi, \Delta, 0)$, i.e., we are simplifying formula $\phi$ by assuming that
domain $\Delta$ is empty. Informally, if there is a predicate $P$ in $\phi$ that
has nothing to do with domain $\Delta$, the role of smoothing is to preserve all
occurrences of $P$ even if all clauses with $P$ become vacuously satisfied. We
note that the approach presented in this section is far from unique and explain
it via an example below.

\begin{example}\label{example:basecasesmoothing}
  Let $\phi$ be
  \begin{align}
    \forall x \in \Delta\text{. }\forall y, z \in \Gamma&\text{. }Q(x) \lor P(y, z)\label[clause]{eq:example1}\\
    \forall y, z \in \Gamma'&\text{. }P(y, z)\label[clause]{eq:example2},
  \end{align}
  where $\Gamma' \subseteq \Gamma$ is a domain introduced by a compilation rule.
  Note that, as a relation, $P \subseteq \Gamma \times \Gamma$.

  Let us reason by hand about the model count of $\phi$ when
  $\Delta = \emptyset$. Predicate $Q$ can only take one value, i.e.,
  $Q = \emptyset$. The value of $P$ is fixed over $\Gamma' \times \Gamma'$ by
  \cref{eq:example2} but is allowed to vary freely over
  $(\Gamma \times \Gamma) \setminus (\Gamma' \times \Gamma')$ since
  \cref{eq:example1} is vacuously satisfied by all structures. Hence, the right
  model count should be $2^{|\Gamma|^2 - |\Gamma'|^2}$.

  However, without \cref{line:smoothing2}, \Propagate would simplify $\phi$ to
  $\forall y, z \in \Gamma'\text{. }P(y, z)$. Here, $P$ is defined as a subset
  of $\Gamma' \times \Gamma'$. Clearly, this simplified formula has only one
  model: $\{\, P(y, z) \mid y, z \in \Gamma' \,\}$.

  With \cref{line:smoothing2} included, $\phi$ is transformed to
  \begin{gather*}
    \forall y, z \in \Gamma\text{. }P(y, z) \lor \neg P(y, z)\\
    \forall y, z \in \Gamma'\text{. }P(y, z),
  \end{gather*}
  which retains the correct model count.
\end{example}

Note that the choice of $l$ on \cref{line:smoothing1} of \cref{alg:propagate} is
inconsequential because any choice achieves the same goal: constructing a
tautological clause that retains the literals in $C'$.

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item It seems like I'll need to introduce:
          \begin{itemize}
            \item FCGs
            \item node types (be consistent w.r.t.\ node vs vertex)
            \item their notation
          \end{itemize}
    \item Describe \cref{fig:smoothing1,fig:smoothing2}
    \item Introduce the compilation rules
  \end{itemize}
}

\begin{figure}
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\DR(c \in \Delta)$,name=dr,ellipse,draw,label={right:\textcolor{blue}{3. $\{\, \forall x \in \Delta\text{. }P(x) \,\}$}}
    [$\land$,ellipse,draw,label={left:\textcolor{blue}{2. $\{\, P(c) \,\}$}},label={right:\textcolor{blue}{6. $\{\, P(c), \forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \,\}$}}
    [$P(c)$,rectangle,draw,fill=green!20,label={272:\textcolor{blue}{1. $\{\, P(c) \,\}$}}]
    [$\CR(\Delta' \gets \Delta \setminus \{\, c \,\})$,ellipse,draw,label={87:\textcolor{blue}{5. $\{\, \forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \,\}$}}
    [$\Reff(\Delta \mapsto \Delta')$,name=ref,ellipse,draw,label={below:\textcolor{blue}{4. $\{\, \forall x \in \Delta'\text{. }P(x) \,\}$}}]
    ]
    ]
    ]
  \end{forest}
  \caption{An example based on bijection counting (but simplified a lot). The
    arc from the $\Reff$ node to the $\DR$ node is
    omitted.}\label{fig:smoothing1}
\end{figure}

\begin{figure}
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\DR(c \in \Delta)$,name=dr,ellipse,draw
    [$\land$,ellipse,draw,dashed,edge=dashed
    [$P(c)$,rectangle,draw,fill=green!20]
    [$\forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \lor \neg P(x)$,rectangle,dashed,edge=dashed,draw,fill=blue!20]
    ]
    ]
  \end{forest}
  \caption{An artificial example of a situation when smoothing nodes need to be
    added below a $\DR$ node. The dashed part is added during
    smoothing}\label{fig:smoothing2}
\end{figure}

Smoothing is a two-step process. First, atoms that are still accounted for in
the circuit are propagated upwards. Then, at vertices of certain types, missing
atoms are detected and additional sinks are created to account for them. If left
unchanged, the first step of this process would result in an infinite loop
whenever a cycle is encountered.

@@

\begin{description}
  \item[\texttt{Ref}.] During smoothing, when unit clauses are propagated in the
        opposite direction of the FCG arcs (i.e., `upwards'), when visiting a
        \texttt{Ref} node, these clauses are translated using the domain map of
        the \texttt{Ref} node. For example, if the domain map include
        $\Delta \mapsto \Delta'$, and the unit clause mentions $\Delta$, then
        replace it by $\Delta'$.
  \item[Constraint removal.] Do reverse constraint removal. Assuming that
        domain $\Delta$ with constraints $x \ne c$ (for some constant
        $c \in \Delta$) were replaced with domain $\Delta'$, replace each
        $\forall x \in \Delta'\text{. }\phi(x)$ with
        $\forall x \in \Delta\text{. }X \ne c \Rightarrow \phi(x)$.
  \item[Domain recursion.] Suppose the domain recursion node introduces constant
        $c \in \Delta$. For each unit clause received from the child node,
        replace each occurrence of $\phi(c)$ or $\forall x \in \Delta\text{.
        } x \ne c \Rightarrow \phi(x)$ with $\forall x \in \Delta\text{.
        }\phi(x)$. This can be seen as a claim about what ground atoms the
        domain recursion node \emph{should} cover (or a temporary assumption).
        If the relevant subgraph indeed covers those ground atoms, Stage~2 will
        do nothing. Otherwise, smoothing nodes will be added below the domain
        recursion node to cover the difference between what was propagated from
        the domain recursion node and what was received from the child node.
\end{description}

@@
